---
title: "Blaney_Module9HW"
author: "Patrick Blaney"
date: "March 11, 2019"
output: pdf_document
---

# Data and package required for completion of assignment
```{r, message=FALSE}
library(multtest)
data("golub")
data("stackloss")
```


# Problem 1

# (a)
```{r, echo=FALSE}
GRO2_index <- grep("GRO2 GRO2 oncogene", golub.gnames[ , 2])
GRO3_index <- grep("GRO3 GRO3 oncogene", golub.gnames[ , 2])

GRO2_data <- golub[GRO2_index, ]
GRO3_data <- golub[GRO3_index, ]
print("The linear correlation coefficient is")
cor(GRO2_data, GRO3_data)

plot(GRO2_data, GRO3_data, xlim = c(-1.6, -0.75), ylim = c(-1.5, -0.5))
```

# (b)
```{r, echo=FALSE}
cor.test(GRO2_data, GRO3_data, conf.level = 0.90)
```
The 90% Confidence Interval is 0.670 to 0.878. The sample estimates given is identical to part (a). Furthermore, with a 
very small p-value of 2.2e-09, the null hypothesis of no correlation can be rejected.

# (c)
```{r, echo=FALSE}
n_boot_1 <- 2000
boot_cor_matrix <- matrix(NA, nrow = n_boot_1, ncol = 1)
GRO_data <- cbind(GRO2_data, GRO3_data)
for(i in 1:n_boot_1) {
  resample_data <- GRO_data[sample(1:nrow(GRO_data), replace = TRUE), ]
  boot_cor_matrix[i, ] <- cor(resample_data[,1], resample_data[,2])
}
quantile(boot_cor_matrix[,1], c(0.05, 0.95))
```


# Problem 2

# (a)
```{r, echo=FALSE}
Zyxin_data <- golub[2124, ]
Zyxin_cors <- apply(golub, 1, function(x) cor(Zyxin_data, x))
print("# of genes with highly negative correlations coefficient to Zyxin")
sum(Zyxin_cors < -0.5)
```

# (b)
```{r, echo=FALSE}
sorted_Zyxin_cors <- order(Zyxin_cors, decreasing = FALSE)
#head(sorted_Zyxin_cors, n = 5)
golub.gnames[394, 2]
Zyxin_cors[394]
golub.gnames[717, 2]
Zyxin_cors[717]
golub.gnames[2489, 2]
Zyxin_cors[2489]
golub.gnames[849, 2]
Zyxin_cors[849]
golub.gnames[1145, 2]
Zyxin_cors[1145]
```

# (c)
```{r, echo=FALSE}
Zyxin_cors_ttest_p_values <- apply(golub, 1, function(x) cor.test(Zyxin_data, x, alternative = "less")$p.value)
Zyxin_cors_ttest_adj_p_values <- p.adjust(Zyxin_cors_ttest_p_values, method = "fdr")
print("# of genes negatively correlated with Zyxin at an FDR adjusted level of 0.05")
sum(Zyxin_cors_ttest_adj_p_values < 0.05)
```


# Problem 3

# (a)
```{r, echo=FALSE}
GRO3 <- golub[2715,]
GRO2 <- golub[2714,]

GRO3_on_GRO2_reg <- lm(GRO3 ~ GRO2)
summary(GRO3_on_GRO2_reg)
```
There is indeed a statistically significant linear relationship between the two genes. The associated p-value for
the correlation coefficient are very small at 2.20e-09. This means we can reject the null hypothesis that
the correlation coefficient is zero.
63.46% of GRO3 GRO3 oncogene expression's variation can be explained by this regression on GRO2 GRO2 oncogene
expression

# (b)
```{r, echo=FALSE}
confint(GRO3_on_GRO2_reg, level = 0.95)
```
The slope is less than 0.5 at the a = 0.05 level. The confidence interval above shows slope with 95% CI (0.266, 0.450)

# (c)
```{r, echo=FALSE}
predict(GRO3_on_GRO2_reg, newdata = data.frame(GRO2=0), interval = "prediction", level = 0.8)
```
The 80% PI when GRO2 GRO2 oncogene is not expressed is (-1.268, -0.418)

# (d)
```{r, echo=FALSE}
shapiro.test(residuals(GRO3_on_GRO2_reg))
plot(GRO3_on_GRO2_reg, which=2)
plot(GRO3_on_GRO2_reg, which=1)
log_GRO3_on_GRO2_reg <- lm(log(GRO3) ~ log(GRO2))
shapiro.test(residuals(log_GRO3_on_GRO2_reg))
```
The Shapiro-Wilk test gave a p-value of 0.0753, which is very close to the standard cutoff but the visual
provided by the qq-plot suggest that the noise is normally distributed but only by a slight margin.
The residuals vs fitted plot appears to show a concentrated linear pattern in the majority of means.
This would suggest the variances are not significantly increasing with the fitted values, so the
homoscedasticity assumption doesn't appear violated.
Another Shapiro-Wilk test after performing a log transformation of the data before regression shows
a p-value of 0.912 which would help reaffirm that the noise is normally distributed.
Thus, we can trust the statistical inferences from the original regression fit. 


# Problem 4

# (a)
```{r, echo=FALSE}
stackloss_regression <- lm(stack.loss ~ Air.Flow + Water.Temp + Acid.Conc., data = stackloss)
summary(stackloss_regression)
```
The fitted regression equation: stack.loss = -39.920 + 0.716 Air.Flow + 1.295 Water.Temp - 0.152 Acid.Conc.

# (b)
Based on the regression, only air flow and water temperature have a statistically significant effect on stack loss.
This derived from their high correaltion coefficients 0.716 and 1.295, respectively, with associated 
p-values of 0.00375 and 5.8e-05. Acid concentration has a very minimal coefficient of -0.152 with a high
p-value of 0.344.
These variables in this regression explain 91.36% of total variation in stack loss.

# (c)
```{r, echo=FALSE}
predict(stackloss_regression,
        newdata = data.frame(Air.Flow=60, Water.Temp=20, Acid.Conc.=90),
        interval = "confidence",
        level = 0.9
        )
print("90% CI of stackloss when Air.Flow=60, Water.Temp=20 and Acid.Conc.=90 is 13.501 to 16.966")
predict(stackloss_regression,
        newdata = data.frame(Air.Flow=60, Water.Temp=20, Acid.Conc.=90),
        interval = "prediction",
        level = 0.9
        )
print("90% PI of stackloss when Air.Flow=60, Water.Temp=20 and Acid.Conc.=90 is 9.331 to 21.136")
```

