---
title: "Blaney_Module3HW"
author: "Patrick Blaney"
date: "January 28, 2019"
output: pdf_document
---

# Problem 1

```{r, echo=FALSE}
print("b")
```


# Problem 2

```{r, echo=FALSE}
x1_range <- c(0, 1, 2, 3, 4)
x1_func <- function(x) ((2 ^ x) / factorial(x)) * exp(-2)
x1_pdf <- function(x) x1_func(x) * (x %in% x1_range)
print("P(X = 1) is")
x1_pdf(1)

print("P(-2 < X <4) is")
sum(x1_pdf(x1_range) * (-2 < x1_range & x1_range < 4))
```


# Problem 3

```{r, echo=FALSE}
print("n = 3, p = 0.25")
```


# Problem 4

```{r, echo=FALSE}
print("P(Y <= 2 ) is")
pbinom(2, size = 3, prob = 0.25)

y1_range <- (0:3)
print("E(Y) is")
e_y1 <- sum(y1_range * dbinom(y1_range, size = 3, prob = 0.25))
e_y1

print("Var(Y) is")
var_y1 <- sum((y1_range - e_y1) ^ 2 * dbinom(y1_range, size = 3, prob = 0.25) )
var_y1
```


# Problem 5

```{r, echo=FALSE}
m2 <- 3
x2_range <- (2:3)
x2_chi_prob <- sum(dchisq(x2_range, m2))
print("P(1 < X < 4) is")
x2_chi_prob

print("E(X) is")
e_x2 <- m2
e_x2

print("Var(X) is")
var_x2 <- 2 * m2
var_x2

x2_chi_monte_carlo <- rchisq(n = 100000, m2)
chi_monte_carlo_est <- mean((1 < x2_chi_monte_carlo) & (x2_chi_monte_carlo < 4))
print("Monte Carlo estimate is")
chi_monte_carlo_est
```
The Monte Carlo estimate agrees with my answer above to a reasonable degree considering only two data points were used for
the original calculation


# Problem 6

```{r, echo=FALSE}
m3 <- 5
e_x3 <- m3
var_x3 <- 2 * e_x3 

print("E(Y) is")
e_y3 <- 4 * e_x3 - 10
e_y3

print("Var(Y) is")
var_y3 <- 4 ^ 2 * var_x3
var_y3
```
Y does not follow a Chi-square distribution with a degree of freedom m = 10 as Var(Y) is more than 2*m


# Problem 7

# (a)
```{r, echo=FALSE}
zyxin_func <- function(x) dnorm(x, mean = 1.6, sd = 0.4)
integrate(zyxin_func, lower = 1, upper = 1.6)
```

# (b)
```{r, echo=FALSE}
norm_monte_carlo <- rnorm(500000, mean = 1.6, sd = 0.4)
norm_monte_carlo_est <- mean((1 < norm_monte_carlo) & (norm_monte_carlo < 1.6))
norm_monte_carlo_est
```

# (c)
Binomial distribution to be utilized.
P(Y >= 2) = 1 - P(Y <= 1)
```{r, echo=FALSE}
n <- 5
p <- integrate(zyxin_func, lower = 1, upper = 1.6)$value
1 - pbinom(1, size = n, prob = p)
```


# Problem 8

# (a)
```{r, echo=FALSE}
f1_m <- 2
f1_n <- 5
f1_dis_fun <- function(x) df(x, f1_m, f1_n)
f1_dis_mean_func <- function(x) x * f1_dis_fun(x)
f1_mean <- integrate(f1_dis_mean_func, lower = 0, upper = Inf)$value
f1_dis_var_func <- function(x) (x - f1_mean)^2 * f1_dis_fun(x)
f1_var <- integrate(f1_dis_var_func, lower = 0, upper = Inf)$value

f2_m <- 10
f2_n <- 5
f2_dis_fun <- function(x) df(x, f2_m, f2_n)
f2_dis_mean_func <- function(x) x * f2_dis_fun(x)
f2_mean <- integrate(f2_dis_mean_func, lower = 0, upper = Inf)$value
f2_dis_var_func <- function(x) (x - f2_mean)^2 * f2_dis_fun(x)
f2_var <- integrate(f2_dis_var_func, lower = 0, upper = Inf)$value
```

# (b)
```{r, echo=FALSE}
f1_direct_mean <- (f1_n / (f1_n - 2))
f1_direct_var <- ((2 * f1_n ^ 2) * (f1_m + f1_n - 2)) / (f1_m * (f1_n - 2) ^ 2 * (f1_n - 4))
print("X~F(m=2,n=5) direct mean is")
f1_direct_mean
print("X~F(m=2,n=5) direct variance is")
f1_direct_var

f2_direct_mean <- (f2_n / (f2_n - 2))
f2_direct_var <- ((2 * f2_n ^ 2) * (f2_m + f2_n - 2)) / (f2_m * (f2_n - 2) ^ 2 * (f2_n - 4))
print("X~F(m=10,n=5) direct mean is")
f2_direct_mean
print("X~F(m=10,n=5) direct variance is")
f2_direct_var
```

# (c)
Both sets of means and variances are identical when either calculating using density function or direct formula
```{r, echo=FALSE}
print("X~F(m=2,n=5) mean and variance using density function:")
f1_mean
f1_var

print("X~F(m=2,n=5) mean and variance using direct formula:")
f1_direct_mean
f1_direct_var

print("X~F(m=10,n=5) mean and variance using density function:")
f2_mean
f2_var

print("X~F(m=10,n=5) mean and variance using direct formula:")
f2_direct_mean
f2_direct_var

```

