---
title: "Blaney_Module12HW"
author: "Patrick Blaney"
date: "April 8, 2019"
output: pdf_document
---

# Data and packages required for completion of assignment
```{r, message=FALSE}
library(ALL)
data(ALL)
library(scatterplot3d)
library(annotate)
```

# Problem 1

# (a)
```{r}
ALL.fac <- as.numeric(ALL$BT %in% c("B","B1","B2","B3","B4")) + 1
```

# (b)
```{r, echo=FALSE}
par(mfrow = c(1, 3))
for(i in 1:3) {
  hist(exprs(ALL[i,]), freq = FALSE, main = "", xlab = rownames(exprs(ALL[i,])))
  lines(density(exprs(ALL[i,])), col = "red")
}
```

# (c)
```{r, echo=FALSE}
first.five.genes <- cbind(as.vector(exprs(ALL[1,])),
                          as.vector(exprs(ALL[2,])),
                          as.vector(exprs(ALL[3,])),
                          as.vector(exprs(ALL[4,])),
                          as.vector(exprs(ALL[5,]))
                          )
pairs(first.five.genes, labels = rownames(ALL[1:5,]))
```

# (d)
```{r, echo=FALSE}
gene.index.39317_at <- grep("39317_at", x = rownames(ALL))
gene.data.39317_at <- as.vector(exprs(ALL[gene.index.39317_at,]))
gene.index.32649_at <- grep("32649_at", x = rownames(ALL))
gene.data.32649_at <- as.vector(exprs(ALL[gene.index.32649_at,]))
gene.index.481_at <- grep("^481_at", x = rownames(ALL), perl = TRUE)
gene.data.481_at <- as.vector(exprs(ALL[gene.index.481_at,]))
genes.of.interest <- cbind(gene.data.39317_at,
                           gene.data.32649_at,
                           gene.data.481_at
                           )
scatterplot3d(genes.of.interest, color = ALL.fac)
```
The two patient groups are fairly well distinguishable from this 3D scatterplot. It might be more visable if the
x, y, and z axis are reorganized.

# (e)
```{r, echo=FALSE}
kmeans.2.cl <- kmeans(genes.of.interest, centers = 2)
table(kmeans.2.cl$cluster)
kmeans.3.cl <- kmeans(genes.of.interest, centers = 3)
table(kmeans.3.cl$cluster)
```
The first method of K-means clustering with 2 centers appears to be on the right track but still has a
relatively high amount of misclassifications at 19. At first glance the K-means clustering with 3 centers
wouldn't seem appropriate but it was actually very accurate. It clustered both "main" patient groups
without any misclassification and further broke the B-cell patient group into two separate subgroups.

# (f)
```{r, echo=FALSE}
ALL.pca <- prcomp(exprs(ALL))
summary(ALL.pca)
```
The first prinicpal component explains the vast majority of the variance in the ALL dataset at 93.59%.
The second prinicpal component explains less than one percent of the rest of the variance at 0.95% and each
principal component after this is even smaller.

# (g)
```{r, echo=FALSE}
biplot(ALL.pca, cex = 0.25, xlim = c(-0.041, 0.018), ylim = c(-0.039, 0.04))
ALL.pca$rotation[,1]
mean(ALL.pca$rotation[,1])
```
Since all the red arrows appear to be nearly identical length, this suggests that PC1 represents the average of
all 128 patients. This is further corroborated by looking at all the loadings for PC1 and then the average of all
the loadings. These values are all very similar confirming the suggestion made by the biplot.

# (h)
```{r, echo=FALSE}
#head(order(ALL.pca$x[,2], decreasing = TRUE), n = 3)
print("The three genes with the largest PC2 values are")
rownames(exprs(ALL[12047]))
rownames(exprs(ALL[12582]))
rownames(exprs(ALL[8095]))
#tail(order(ALL.pca$x[,2], decreasing = TRUE), n = 3)
print("The three genes with the smallest PC2 values are")
rownames(exprs(ALL[4721]))
rownames(exprs(ALL[2673]))
rownames(exprs(ALL[9406]))
```

# (i)
```{r, echo=FALSE, message=FALSE}
library('hgu95av2.db')
pc2.largest.gene.manf <- rownames(exprs(ALL[12047]))
pc2.smallest.gene.manf <- rownames(exprs(ALL[4721]))
get(pc2.largest.gene.manf, envir = hgu95av2GENENAME)
get(pc2.smallest.gene.manf, envir = hgu95av2GENENAME)
get(pc2.largest.gene.manf, envir = hgu95av2CHR)
get(pc2.smallest.gene.manf, envir = hgu95av2CHR)
```


# Problem 2

# (a)
```{r}
iris.data <- data.frame(iris$Sepal.Length,
                        iris$Sepal.Width,
                        iris$Petal.Length,
                        iris$Petal.Width
                        )
scaled.iris.data <- data.frame(log(iris$Sepal.Length),
                               log(iris$Sepal.Width),
                               log(iris$Petal.Length),
                               log(iris$Petal.Width)
                               )
```

# (b)
```{r, echo=FALSE}
print("Correlation between unscaled sepal length and sepal width")
cor(iris.data$iris.Sepal.Length, iris.data$iris.Sepal.Width)
print("Correlation between scaled sepal length and sepal width")
cor(scaled.iris.data$log.iris.Sepal.Length., scaled.iris.data$log.iris.Sepal.Width.)

print("Correlation between unscaled petal length and petal width")
cor(iris.data$iris.Petal.Length, iris.data$iris.Petal.Width)
print("Correlation between scaled petal length and petal width")
cor(scaled.iris.data$log.iris.Petal.Length., scaled.iris.data$log.iris.Petal.Width.)

print("Correlation between unscaled sepal length and petal width")
cor(iris.data$iris.Sepal.Length, iris.data$iris.Petal.Width)
print("Correlation between scaled sepal length and petal width")
cor(scaled.iris.data$log.iris.Sepal.Length., scaled.iris.data$log.iris.Petal.Width.)
```

# (c)
```{r}
#dist(cbind(scaled.iris.data$log.iris.Sepal.Length., scaled.iris.data$log.iris.Sepal.Width.), method = "euclidean") ^ 2
```

# (d)
```{r, echo=FALSE}
unscaled.iris.pca <- prcomp(iris.data, scale. = FALSE)
unscaled.iris.pca
scaled.iris.pca <- prcomp(scaled.iris.data, scale. = FALSE)
scaled.iris.pca
```
While the two principal component analyses are fairly similar they are not the identical.

# (e)
```{r, echo=FALSE}
summary(scaled.iris.pca)
summary(unscaled.iris.pca)
```
For the scaled PCA, PC1 and PC2 explain 97.02% and 1.413%, respecitively.
For the unscaled PCA, PC1 and PC2 explain 92.46% and 5.307%, respectively.

# (f)
```{r, echo=FALSE}
quantile(scaled.iris.pca$x[,2], c(0.05, 0.95))
```

