---
title: "Blaney_Module10HW"
author: "Patrick Blaney"
date: "March 26, 2019"
output: pdf_document
---

# Problem 1

# (a)
```{r, echo=FALSE, message=FALSE}
library('ArrayExpress')
library('affy')
getAE('E-MEXP-1551', path = 'yeast/', type = 'full')
yeast.raw <- ReadAffy(celfile.path = 'yeast/')
yeast.bgc.norm.pmc <- expresso(yeast.raw,
                               bgcorrect.method = 'mas',
                               normalize.method = 'quantiles',
                               pmcorrect.method = 'pmonly',
                               summary.method = 'medianpolish'
                               )
```

# (b)
```{r, echo=FALSE}
yeast.exprs <- exprs(yeast.bgc.norm.pmc)
yeast.exprs.first.5 <- yeast.exprs[1:5,]
yeast.exprs.first.5.means <- apply(yeast.exprs.first.5, 1, mean)
yeast.exprs.first.5.means
```

# (c)
```{r, echo=FALSE}
dim(yeast.bgc.norm.pmc)
```
There are a total of 10928 genes across 30 samples in this preprocessed expression data set.


# Problem 2

# (a)
```{r, echo=FALSE, message=FALSE}
annotation(yeast.raw)
library('yeast2.db')
```
The annotation package for the yeast data set is 'yeast2'

# (b)
```{r, echo=FALSE}
library('XML')
library('annotate')
go.num.1769308_at <- get('1769308_at', env = yeast2GO)
mf.go.num.1769308_at <- getOntology(go.num.1769308_at, 'MF')
mf.go.num.1769308_at
length(mf.go.num.1769308_at)
```
There are a total of 6 GO numbers of the 1769308_at gene related to molecular function.

# (c)
```{r, echo=FALSE}
go.parents.1769308_at <- getGOParents(mf.go.num.1769308_at)
go.parents.list.1769308_at <- sapply(go.parents.1769308_at, function(x) x$Parents)
go.parents.num.1769308_at <- unique(unlist(go.parents.list.1769308_at))
go.parents.num.1769308_at
length(go.parents.num.1769308_at)
```
The 1769308_at gene molecular function related GO numbers have 4 parent GO numbers.

# (d)
```{r, echo=FALSE}
go.children.1769308_at <- getGOChildren(mf.go.num.1769308_at)
go.children.list.1769308_at <- sapply(go.children.1769308_at, function(x) x$Children)
go.children.num.1769308_at <- unique(unlist(go.children.list.1769308_at))
go.children.num.1769308_at
length(go.children.num.1769308_at)
```
The 1769308_at gene molecular function related GO numbers have 132 children GO numbers.


# Problem 3

# (a)
```{r, message=FALSE}
library("ALL")
data('ALL')
library('genefilter')
b2.b3.stage.patients <- factor(ALL$BT %in% c("B2","B3"))
wilcoxon.filter <- function(x) {
  (wilcox.test(x ~ b2.b3.stage.patients, exact = FALSE, paired = FALSE)$p.value < 0.001)
}
welsh.filter <- function(x) {
  (t.test(x ~ b2.b3.stage.patients)$p.value < 0.001)
}
wilcoxon.filtered.exprs <- genefilter(exprs(ALL), wilcoxon.filter)
welsh.filtered.exprs <- genefilter(exprs(ALL), welsh.filter)
```

# (b)
```{r, echo=FALSE, message=FALSE}
library('limma')
all.filtered.exprs <- apply(cbind(wilcoxon.filtered.exprs, welsh.filtered.exprs), 2, as.integer)
filter.counts <- vennCounts(all.filtered.exprs, include = 'both')
vennDiagram(filter.counts, names = c('Wilcoxon', 'Welsh'))
```

# (c)
79 genes pass only the Wilcoxon filter at a level of 0.001. A total of 648 genes pass both the Wilcoxon
and Welsh filters.

# (d)
```{r, echo=FALSE, message=FALSE}
annotation(ALL)
library('hgu95av2.db')
gene.manuf.name <- rownames(exprs(ALL))
find.genename <- function(x) get(x, envir = hgu95av2GENENAME)
gene.names.vec <- rep(NA, length(gene.manuf.name))
for(i in 1:length(gene.manuf.name)) {
  gene.names.vec[i] <- find.genename(gene.manuf.name[i])
}
all.oncogene <- unique(gene.names.vec[grep('oncogene', gene.names.vec)])
all.oncogene
```
The annotation package for the ALL data set is hgu95av2.db.

# Problem 4

# (a)
```{r}
b1.b2.b3.stage.patients <- ALL[,which(ALL$BT %in% c("B1","B2","B3"))]
```

# (b)
```{r, echo=FALSE}
anova.design.model <- model.matrix(~ 0 + factor(b1.b2.b3.stage.patients$BT))
colnames(anova.design.model) <- c("B1","B2","B3")
linear.model.fit <- lmFit(b1.b2.b3.stage.patients, anova.design.model)
linear.model.fit.bayes.adj <- eBayes(linear.model.fit)
topTable(linear.model.fit.bayes.adj, coef = 3, number = 5)
```

# (c)
```{r, echo=FALSE}
contrast.model <- makeContrasts(B1-B2, B2-B3, levels = factor(b1.b2.b3.stage.patients$BT))
contrast.model.fit <- contrasts.fit(linear.model.fit, contrast.model)
contrast.model.fit.bayes.adj <- eBayes(contrast.model.fit)
topTable(contrast.model.fit.bayes.adj, number = 5, adjust.method = 'fdr')
```

