---
title: "Blaney_Module4HW"
author: "Patrick Blaney"
date: "January 30, 2019"
output: pdf_document
---

# Problem 1

# (a)
```{r, echo=FALSE}
mean_1 <- 5
sd_1 <- 3
n_1 <- 5

ex_1 <- mean_1
print("E(X) is")
ex_1

sdx_1 <- sd_1 / sqrt(n_1)
print("sd(X) is")
sdx_1
```

# (b)
```{r, echo=FALSE}
print("Estimate of P(2<X<5.1) is")
pnorm(5.1, mean = ex_1, sd = sdx_1) - pnorm(2, mean = ex_1, sd = sdx_1)
```


# Problem 2

```{r, echo=FALSE}
size_2 <- 20
p_2 <- 0.7
n_2 <- 100
mean_2 <- size_2 * p_2
var_2 <- (size_2 * p_2) * (1 - p_2)

print("P(Y > 15) is")
1 - pnorm(15, mean = mean_2, sd = sqrt(var_2) / sqrt(n_2))
```


# Problem 3

```{r, echo=FALSE, message=FALSE}
require(mvtnorm)
x_mean_3 <- 9
x_var_3 <- 3
y_mean_3 <- 10
y_var_3 <- 5
covar_3 <- 2
n_3 <- 50

distr_3 <- rmvnorm(n_3, mean = c(x_mean_3, y_mean_3), sigma = matrix(c(x_var_3, covar_3, covar_3, x_var_3), nrow = 2))
x_samples_3 <- distr_3[,1]
y_samples_3 <- distr_3[,2]
prob_3 <- mean(y_samples_3 - (x_samples_3 + 0.5)) + c(-1,1) * sqrt(var(y_samples_3 - (x_samples_3 + 0.5)) / n_3)
print("The approximate probability of second gene exceeding sample mean")
print("of first gene by more than 0.5 with 95% interaval is")
prob_3
```


# Problem 4

In order to achieve accurate Monte Carlo simulation, a random sample size of 100,000 will be utlized.
```{r, echo=FALSE}
x_chisq_4 <- rchisq(100000, df = 10)
x_gamma_4 <- rgamma(100000, shape = 1, rate = 2)
x_t_4 <- rt(100000, df = 3)

y_4 <- (sqrt(x_chisq_4) * x_gamma_4) + 4 * (x_t_4) ^ 2
y_mean_4 <- mean(y_4)
print("The mean of Y using Monte Carlo simulation is")
y_mean_4
```
Considering that the formulas for determining the mean of random variables used in more complex equations are out of scope, 
the Monte Carlo estimation is sufficient. This is reenforced even more when you take the built in accuracy of a Monte Carlo 
simulation. The approximate value given by this simulation will vary by about +-0.5 from 13.2. The simulation could only be 
refined more by generating multiple sets of random samples and taking the mean of all those values.


# Problem 5

For the standard normal distribution, mean = 0 and standard deviation = 1,
sample size n = 1000 for determining maxima.

```{r, echo=FALSE}
mean_5 <- 0
sd_5 <- 1
n_5 <- 1000
an <- sqrt(2*log(n_5)) - 0.5*(log(log(n_5))+log(4*pi))*(2*log(n_5))^(-1/2)
bn <- (2*log(n_5))^(-1/2)
ev_dis_func <- function(x) exp(-x) * exp(-(exp(-x)))

norm_maxima_matrix <- matrix(rnorm(n_5 * 1000, mean = mean_5, sd = sd_5), nrow = 1000)
norm_maximas <- (apply(norm_maxima_matrix, 2, max) - an) / bn

hist_plot_5 <- hist(norm_maximas, freq = FALSE, ylim = c(0, 0.4), xlim = c(-3, 6), 
                    main = "Distribution of Normalized Maximas",
                    xlab = "Normalized Maximas")
curve(ev_dis_func(x), col = "red", add = T)
curve(dnorm(x, mean = 0, sd = 1), col = "blue", add = T)
```

Based on the plot of the two curves, red being the extreme value function and blue being the standard normal distribution, the steps
of maxima normalization will approach a high level of accuracy for high sample sizes. The histogram clearly shows that the range of
maximas after normalization are approaching these ideal curves. 

