---
title: "Blaney_Module8HW"
author: "Patrick Blaney"
date: "March 1, 2019"
output: pdf_document
---

# Data and package required for completion of assignment
```{r, message=FALSE}
library(ALL)
data(ALL)
library(lmtest)
```


# Problem 1

# (a)
```{r, echo=FALSE}
ALL_Bstages_1 <- ALL[, ALL$BT %in% c("B", "B1", "B2", "B3", "B4")]
gene_exp_109_at <- exprs(ALL_Bstages_1)["109_at", ]
anova(lm(gene_exp_109_at ~ ALL_Bstages_1$BT))
```
The p-value from this one-way ANOVA table is 0.01082 which is low enough to reject the null hypothesis that means
are all equal. Thus, it is concluded that the disease stages, B, B1, B2, B3, and B4 affect the mean gene expression value.

# (b)
```{r, echo=FALSE}
print("Summary of linear model without intercept")
summary(lm(gene_exp_109_at ~ ALL_Bstages_1$BT -1))

print("Summary of linear model with intercept")
summary(lm(gene_exp_109_at ~ ALL_Bstages_1$BT))
```
From these summary statistics, the mean gene expression value among the B3 patients is 6.68533. From table one, this
value is estimated for B3 as 6.68533. This value is confirmed in table two, as calulated by 6.8102-0.1249=6.6853.

# (c)
```{r, echo=FALSE}
pairwise.t.test(gene_exp_109_at, ALL_Bstages_1$BT)
```
Based on the compairative pairwise t-test table, only B2's mean showed any difference from B but with p-value of
only 0.52 this difference cannot be confirmed

# (d)
```{r, echo=FALSE}
pairwise.t.test(gene_exp_109_at, ALL_Bstages_1$BT, p.adjust.method = 'fdr')
```
Based on the compairative pairwise t-test table with the p-values adjusted with the FDR method, only the mean gene
expression value of B4 is different from B2. This is confirmed witha adjusted p-value of 0.01 which passes the 0.05 level test

# (e)
```{r, echo=FALSE}
shapiro.test(residuals(lm(gene_exp_109_at ~ ALL_Bstages_1$BT)))
bptest(lm(gene_exp_109_at ~ ALL_Bstages_1$BT))
```
With the Shapiro-Wilk test for normality giving a p-value of 0.1177, the null hypothesis of normally
distributed residuals cannot be rejected. Therefore, the group gene expression residuals are normal.
With the Breusch-Pagan test for homoscedasticity giving a p-value of 0.8352, the null hypothesis of 
equal variances cannot be rejected. Therefore, the group gene expression residuals are homoscedastic.
With both assumptions being confirmed, there is no need for a more robust test.


# Problem 2

# (a)
```{r, echo=FALSE}
ALL_Bstages_2 <- ALL[, ALL$BT %in% c("B", "B1", "B2", "B3", "B4")]
gene_exp_ALL <- exprs(ALL_Bstages_2)
kw_tests <- apply(gene_exp_ALL, 1, function(x) kruskal.test(x ~ ALL_Bstages_2$BT)$p.value)
kw_tests_fdr_adj <- p.adjust(kw_tests, method = 'fdr')
print("# of genes within the groups that pass the 0.05 FDR adjusted level are")
sum(kw_tests_fdr_adj < 0.05)
```
Since these genes passed the FDR adjusted p-value level, the null hypothesis of equal distributions of gene
expression values can be rejected.

# (b)
```{r, echo=FALSE}
sorted_kw_adj_p <- order(kw_tests_fdr_adj, decreasing = FALSE)
print("Probe names with smallest FDR adjusted p-value")
kw_tests_fdr_adj[419]
kw_tests_fdr_adj[8636]
kw_tests_fdr_adj[10365]
kw_tests_fdr_adj[964]
kw_tests_fdr_adj[10251]
```


# Problem 3

# (a)
```{r, echo=FALSE}
ALL_Bstages_sex <- ALL[ , which(ALL$BT %in% c("B1", "B2", "B3", "B4") & ALL$sex %in% c("M", "F"))]
gene_exp_38555_at <- exprs(ALL_Bstages_sex)["38555_at", ]
Bstages <- ALL_Bstages_sex$BT
sex <- ALL_Bstages_sex$sex
anova(lm(gene_exp_38555_at ~ Bstages * sex))
```
Based on the ANOVA table, the p-value for the disease stages is very small showing this is a factor.
However, sex is not a factor that affects gene expression and there also no statistically significant interaction
based on their respective p-values 0.7851 and 0.9095

# (b)
```{r, echo=FALSE}
shapiro.test(residuals(lm(gene_exp_38555_at ~ Bstages * sex)))
bptest(lm(gene_exp_38555_at ~ Bstages * sex))
```
Based on the output, the Shapiro-Wilk normality test p-value of 0.03291 indicates the null hypothesis of normality
can be rejected. This violates the assumption of normality and would require a more robust test.
However, the Breusch-Pagan homoscedasticity test p-value of 0.3209 indicates the null hypothesis of equal variances
cannot be rejected. This does not violate the assumption of equal variances.


# Problem 4

# (a)
```{r}
anova_perm_test <- function(gene_data, factor_group) {
  n <- length(gene_data)
  factor_group_means <- by(gene_data, factor_group, mean)
  n_factor_group <- length(factor_group_means)
  obs_stat <- (1 / (n_factor_group - 1)) * sum((factor_group_means - mean(factor_group_means)) ^ 2)
  
  n_perm <- 2000
  perm_stat <- rep(NA, n_perm)
  for(i in 1:n_perm) {
    perm_data <- sample(gene_data, n, replace = FALSE)
    perm_factor_means <- by(perm_data, factor_group, mean)
    n_perm_factor <- length(perm_factor_means)
    perm_stat[i] <- (1 / (n_perm_factor - 1)) * sum((perm_factor_means - mean(perm_factor_means)) ^ 2)
  }
  
  mean(perm_stat >= obs_stat)
}
```

# (b)
```{r, echo=FALSE}
ALL_Bstages_4 <- ALL[, ALL$BT %in% c("B1", "B2", "B3")]
gene_exp_1242_at <- exprs(ALL_Bstages_4)["1242_at", ]
Bstage_factor_group <- ALL_Bstages_4$BT[, drop = TRUE]
anova_perm_test(gene_exp_1242_at, Bstage_factor_group)
```
With a permutated p-value of 0.524, the the null hypothesis that means are all equal cannot be rejected.
