---
title: "Final_Exam"
author: "Patrick Blaney"
date: "April 20, 2019"
output: pdf_document
---

# Problem 1

# (a)
```{r, echo=FALSE}
f.x.1 <- function(x) {2.469862 * x * exp(-x ^ 2)}
f.y.1 <- function(y) {2 * y * exp(-y ^ 2)}

x.range.1 <- c(1:3)
e.x.1 <- sum(x.range.1 * f.x.1(x.range.1))
print("E(X) is")
e.x.1

e.y.1 <- integrate(function(y) y * f.y.1(y), lower = 0, upper = Inf)
print("E(Y) is")
e.y.1$value

var.x.1 <- sum((x.range.1 - e.x.1) ^ 2 * f.x.1(x.range.1))
sd.x.1 <- sqrt(var.x.1)
print("sd(X) is")
sd.x.1

var.y.1 <- integrate(function(y) (y - e.y.1$value)^2 * f.y.1(y), lower = 0, upper = Inf)
sd.y.1 <- sqrt(var.y.1$value)
print("sd(Y) is")
sd.y.1
```

# (b)
```{r, echo=FALSE}
e.x.y.1 <- (2 * e.x.1) - (3 * e.y.1$value)
print("E(2X - 3Y) is")
e.x.y.1

sd.x.y.1 <- (2 * sd.x.1) - (3 * sd.y.1)
print("sd(2X - 3Y) is")
sd.x.y.1
```


# Problem 2
```{r, echo=FALSE}
f.x.2 <- function(x) dnorm(x, mean = 0, sd = 1)
f.y.2 <- function(y) dchisq(y, df = 4)

e.x.2 <- integrate(function(x) x * f.x.2(x), lower = 0, upper = Inf)$value
e.y.2 <- integrate(function(y) y * f.y.2(y), lower = 0, upper = Inf)$value

e.x.y.2 <- (2 * e.x.2 ^ 2) / (e.x.2 ^ 2 + e.y.2)
print("E(2X^2 / X^2 + Y) is 0.08")
```


# Problem 3
Due to the nature of Monte Carlo simulations and the general Rule of Large Numbers, the probability of the
simulation derived empirical coverage being greater than 94% when the true coverage, based on CI formula used,
is 92%, should be very low. The simulation coverage will further approach 92% as the number of simulations is
increased as well.


# Problem 4

# (a)
```{r, message= FALSE, echo=FALSE}
library(multtest)
data("golub")
GRO3.data <- golub[2715,]
MYC.data <- golub[2302,]
GRO3.MYC.diff <- GRO3.data - MYC.data
prop.test(x = sum(GRO3.MYC.diff > 0), n = length(GRO3.MYC.diff), p=0.5, alternative = "two.sided")
```
With a p-value of 0.058, the null hypothesis cannot be rejected. Thus, it cannot be concluded that the median
difference between the expression values of CRO3 gene and MYC gene is less than zero.

# (b)
```{r, echo=FALSE}
binom.test(x = sum(GRO3.MYC.diff > 0), n = length(GRO3.MYC.diff), p = 0.5, alternative = "greater")
```
The 95% one-sided upper CI for the median difference between the expression values of GRO3 gene and MYC gene
is (0.150, 1.000). Additionally, with a p-value of 0.999, the null hypothesis cannot be rejected which confirms
the previous assessment.

# (c)
```{r, echo=FALSE}
n.4 <- length(GRO3.MYC.diff)
n.boot.4 <- 1000
boot.vals <- rep(NA, n.boot.4)
for(i in 1:n.boot.4){
  data.samples <- GRO3.MYC.diff[sample(1:n.4, replace = TRUE)]
  boot.vals[i] <- mean(data.samples)
}
quantile(boot.vals, c(0.025, 0.975))
```


# Problem 5

# (a)
```{r, echo=FALSE}
HPCAH.index <- grep("HPCA Hippocalcin", golub.gnames[,2])
print("HPCA Hippocalcin gene is row number")
HPCAH.index
```

# (b)
```{r, echo=FALSE}
gol.fac <- factor(golub.cl, levels = c(0,1), labels = c("ALL", "AML"))
HPCAH.ALL.data <- golub[HPCAH.index,gol.fac=="ALL"]
HPCAH.ALL.neg.tot <- sum(HPCAH.ALL.data < 0)
print("Proportion of ALL patients that the HPCA Hippocalcin gene is negatively expressed is")
HPCAH.ALL.neg.tot/length(HPCAH.ALL.data)
```

# (c)
The null hypothesis is is H0: mu = 0 and the alternative hypothesis is HA: mu != 0. This will require a one-sided t-test.
```{r, echo=FALSE}
t.test(HPCAH.ALL.data, mu = 0, alternative = "less")
```
With a p-value of 0.01331, the null hypothesis can be rejected. Thus, it can be concluded that the HPCA Hippocalcin
gene is negatively expressed in half of the population. This is also confirmed by the sample estimates mean being 
-0.280 and reaffirms the calculation above that 59.26% of the ALL patients negatively expressed the gene.

# (d)
```{r, echo=FALSE}
HPCAH.AML.data <- golub[HPCAH.index, gol.fac=="AML"]
HPCAH.AML.neg.tot <- sum(HPCAH.AML.data < 0)
prop.test(x = c(HPCAH.ALL.neg.tot, HPCAH.AML.neg.tot),
          n = c(length(HPCAH.ALL.data), length(HPCAH.AML.data)),
          alternative = "two.sided")
```
The 95% CI for the difference of proportions in the ALL group versus the AML group with negatively
expressed HPCA Hippocalcin gene is (-0.348, 0.442)


# Problem 6

# (a)
```{r, message=FALSE}
library(ALL)
data(ALL)
ALL.B.cell.data <- exprs(ALL)[,1:95]
```

# (b)
```{r, echo=FALSE}
library(genefilter)
covar.filter <- cv(0.2)
filtered.ALL.B.cell.genes <- genefilter(ALL.B.cell.data, filterfun(covar.filter))
print("Total number of genes with a coefficient of variance greater than 0.2")
sum(filtered.ALL.B.cell.genes)
```

# (c)
```{r, echo=FALSE}
filtered.ALL.B.cell.data <- ALL.B.cell.data[filtered.ALL.B.cell.genes,]
hclust.filtered.ALL.B.cell <- hclust(dist(t(filtered.ALL.B.cell.data), method = "euclidean"), method = "ward.D2")
filtered.ALL.B.cell.clusters <- cutree(hclust.filtered.ALL.B.cell, k = 4)
ALL.B.cell.classes <- as.character(ALL$BT[1:95])
table(ALL.B.cell.classes, filtered.ALL.B.cell.clusters)
ALL.mol.bio.classes <- as.character(ALL$mol.biol[1:95])
table(ALL.mol.bio.classes, filtered.ALL.B.cell.clusters)
```
The clusters compare fairly well to the B-stages but there is a relatively high misclassification rate, especially
for the smaller true classes like B1 and B4.
On the otherhand, the clusters did quite well for the smaller true classes in the molecular biology types. The
ALL1/AF4, E2A/PBX1, and p15/p16 types were classified without error but the larger groups saw a much harder
time to discern one from the other.

# (d)
```{r, echo=FALSE, message=FALSE}
library(gplots)
color.map.B.cell <- function(B) {
  if(B=="B1") "orange"
  else if(B=="B2") "yellow"
  else if(B=="B3") "purple"
  else if(B=="B4") "lightblue"
  else "gray"
}
B.cell.colors <- unlist(lapply(ALL.B.cell.classes, color.map.B.cell))
heatmap.2(filtered.ALL.B.cell.data,
          col = greenred(75),
          scale = "row",
          dendrogram = "col",
          ColSideColors = B.cell.colors,
          margin = c(3,12), cexRow = 0.5,
          key = FALSE, trace = "none", labRow = NA, labCol = NA)

color.map.mol.bio <- function(M) {
  if(M=="ALL1/AF4") "orange"
  else if(M=="BCR/ABL") "yellow"
  else if(M=="E2A/PBX1") "purple"
  else if(M=="NEG") "lightblue"
  else "gray"
}
mol.bio.colors <- unlist(lapply(ALL.mol.bio.classes, color.map.mol.bio))
heatmap.2(filtered.ALL.B.cell.data,
          col = greenred(75),
          scale = "row",
          dendrogram = "col",
          ColSideColors = mol.bio.colors,
          margin = c(3,12), cexRow = 0.5,
          key = FALSE, trace = "none", labRow = NA, labCol = NA)
```
Based on the heatmaps created and the confusion matrices from the previous part of the question, I would
say that the cluster better reflects the molecular biology types.

# (e)
```{r, echo=FALSE, message=FALSE}
B1234.cell.tot <- sum(ALL.B.cell.classes %in% c("B1", "B2", "B3", "B4"))
new.B.cell.classes <- rep(NA, B1234.cell.tot)
new.B.cell.classes.col.index <- rep(NA, B1234.cell.tot)
count <- 1
for(i in 1:length(ALL.B.cell.classes)){
  if(ALL.B.cell.classes[i]=="B1"){
    new.B.cell.classes[count] <- "B1"
    new.B.cell.classes.col.index[count] <- i
    count <- count + 1
  }
  else if(ALL.B.cell.classes[i]=="B2"){
    new.B.cell.classes[count] <- "B2"
    new.B.cell.classes.col.index[count] <- i
    count <- count + 1
  }
  else if(ALL.B.cell.classes[i]=="B3"){
    new.B.cell.classes[count] <- "B34"
    new.B.cell.classes.col.index[count] <- i
    count <- count + 1
  }
  else if(ALL.B.cell.classes[i]=="B4"){
    new.B.cell.classes[count] <- "B34"
    new.B.cell.classes.col.index[count] <- i
    count <- count + 1
  }
}

new.ALL.B.cell.data <- exprs(ALL)[,new.B.cell.classes.col.index]
library(limma)
design.model <- model.matrix(~0 + factor(new.B.cell.classes))
colnames(design.model) <- c("B1", "B2", "B34")
lin.mod.fit <- lmFit(new.ALL.B.cell.data, design = design.model)
lin.mod.fit <- eBayes(lin.mod.fit)
contrast.model <- makeContrasts(B1-B2,B2-B34, B34-B1, levels = factor(new.B.cell.classes))
contrast.mod.fit <- contrasts.fit(lin.mod.fit, contrasts = contrast.model)
contrast.mod.fit <- eBayes(contrast.mod.fit)

fdr.adj.filter1 <- p.adjust(contrast.mod.fit$p.value[,1], method = "fdr") <0.05
fdr.adj.filter2 <- p.adjust(contrast.mod.fit$p.value[,2], method = "fdr") <0.05
fdr.adj.filter3 <- p.adjust(contrast.mod.fit$p.value[,3], method = "fdr") <0.05
contrast.adj.pvalues <- cbind(fdr.adj.filter1, fdr.adj.filter2, fdr.adj.filter3)
genes.pass.filter <- apply(contrast.adj.pvalues, 1, sum)
print("Number of genes that express differently among B1, B2, and B34 with FDR adjustment")
sum((genes.pass.filter / 3) > 0)
```

# (f)
```{r, echo=FALSE, message=FALSE}
filtered.B1234.data <- new.ALL.B.cell.data[genes.pass.filter,]
library(e1071)
library(rpart)
svm.fit <- svm(new.B.cell.classes ~ t(filtered.B1234.data), type = "C-classification", kernel = "linear")
svm.predict <- predict(svm.fit, t(filtered.B1234.data))
table(svm.predict, new.B.cell.classes)
class.tree <- rpart(new.B.cell.classes ~ ., data = data.frame(t(filtered.B1234.data)))
plot(class.tree, branch = 0, margin = 0.1)
text(class.tree)

n.val <- length(new.B.cell.classes)
mcr.val <- rep(NA, n.val)
#for(i in 1:n.val){
#  svm.fit.test <- svm(t(filtered.B1234.data)[-i],
#                      new.B.cell.classes[-i],
#                     type = "C-classification",
#                     kernel = "linear")
#  svm.predict.test <- predict(svm.fit.test, filtered.B1234.data[i])
#  mcr.val[i] <- mean(svm.predict.test != new.B.cell.classes[i])
#}
```

